BETTING AND BELIEF: PREDICTION MARKETS AND ATTRIBUTION OF CLIMATE CHANGE

### 2016-03-29 prediction_market_wsc.Rnw
Updated SA analysis for PRCC and changed plot ylab to match.
Fixed some typos and did a little wordsmithing.
Moved redundant discussion of historical model runs from DISCUSSION to Historical Climate.

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public \shortcite{Vandenbergh2013f}. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence. The costs of misinformed climate policies are high. If climate change is not human-induced but is believed to be so, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but not recognized as so, the costs of inaction could be devastating. Given that effective climate policies require long-lasting measures to be taken, the importance of a timely and accurate consensus on the issue is manifest.

@@

Bloch, Annan and Bowles \shortciteN{Bloch_2010} have proposed using derivatives markets to reduce the scientific uncertainty in estimating the impact of future climate change. Existing prediction markets (e.g. hypermind.com, betfair.com, and predictit.org) focus on near term events such as elections months away, so it is difficult to extrapolate empirical findings to the climate case. Furthermore, we are interested in investigating the unobservable beliefs of traders. Therefore, we turn to simulation modeling informed by climate and economic theory. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what social and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs. Our work can be extended as part of a computational design process for effective climate prediction markets.

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the comparable predictive power of statistical supervised learning models, which can provide information at a much lower cost than creating and maintaining a market \shortcite{goel_prediction_2010}. Effective climate policy not only requires an accurate consensus on future climate \emph{outcomes\/}. It also requires an accurate consensus on the causal \emph{mechanisms\/} influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented.

@@

Traders bet on global temperature anomalies six years in the future. 
During the six-year period, traders buy and sell futures. 
Every year, traders update their models and forecast of future temperatures based on newly available data. 
At the end of each six-year period, winners collect gains and traders revise beliefs about climate models, based on their ideology and the beliefs of top earners in their social network. 
In this section, we describe the models used to generate future temperature data, agents beliefs about those models, the market procedures, and the social network connecting agents. 
We conclude with details on model dynamics and an overview of the model parameters we experimentally vary.

@@

For climate time-series, we use the annual anomaly of global mean temperature. For years from 1880--2014 we use the GISTEMP global mean land-sea annual temperature anomalies \shortcite{GISTEMP_2016,Hansen_2010} and for years from 2015 onward, we project future climates under two alternative theories: In both theories, changes in global temperature are proportional to changes in a deterministic forcing plus a stochastic noise term. For simplicity, we choose two alternative expressions for the deterministic climate forcing: one, which corresponds to conventional climate science, takes the natural logarithm of the atmospheric carbon dioxide concentration in parts per million \shortcite{Archer_2012}[p.~37], and the other, which corresponds to an alternative theory advocated by many who doubt or reject conventional climate science, takes the total solar irradiance (the brightness of sunlight, in Watts per square meter, at the top of the atmosphere) averaged over the 11-year sunspot cycle \shortcite{Soon_2005}.  Most scientific models of the earth's climate include many forcings, including carbon dioxide, other greenhouse gases, aerosols, total solar irradiance, and more. In these models, the changing $\ce{CO2}$ concentration is, by a large margin, the strongest forcing \shortcite{IPCC_WG1_AR5_2013}[p.~14], so for simplicity we take this as the model. Choosing only one forcing term for each of the competing models simplifies comparison because each has the same number of adjustable parameters.

For $\ce{CO2}$ we used historical emissions through 2005, harmonized with RCP 8.5 representative concentration pathway from 2005 onward \shortcite{rcp_database_2009,Riahi_2011} and for TSI we used the harmonized historical values, with a projection through 2100 from \shortciteN{Velasco_Herrera_2015}, which seems to be the best TSI model.

Warming coefficients for each model were determined by linear regression of historical temperatures from 1880--2014 against the historical values for each model's forcing term. The noise model was determined by fitting an $\ARMA(p,q)$ model to the residuals from the regression, using the \href{http://mc-stan.org}{Stan} probabilistic modeling language and the \href{https://cran.r-project.org/web/packages/rstan/}{\texttt{rstan}} package \shortcite{Stan2016,RStan2016}. We identified the optimal model for the auto-correlated noise term by performing the regression analysis for all combinations of $p, q \in [0,1,2]$ and using the Watanabe-Akaike Information Criterion to select the optimal noise model \shortcite{Vehtari_2015}. In both cases (TSI and $\log\ce{CO2}$), the optimal noise model was $\AR(1)$.

Future climates were generated by applying the warming coefficients to Eq.~\ref{eq:climate}:
\begin{equation}
T_{\text{model}}(t) = \beta_{\text{model}} F_{\text{model}}(t) + \varepsilon,
\label{eq:climate}
\end{equation}
where $T_{\text{model}}(t)$ is the temperature at time $t$, under a given model of what causes warming,
$F_{\text{model}}(t)$ is the forcing (TSI or $\log\ce{CO2}$) at time $t$,
$\beta_{\text{model}}$ is the regression coefficient, and
$\varepsilon$ is an $\ARMA(p,q)$ noise term.
The coefficient $\beta$ and the parameters of the ARMA noise model were fit to the historical data (1880--2014).
An example realization of future climates for both the $\ln(\ce{CO2})$ and TSI models is shown in Fig.~\ref{fig:climateplot}.

@@

Based on traders' sell and buy orders, traders exchange securities following a continuous-double auctions (CDA) procedure (see Model Dynamics below for more details). 
CDAs or some close variants are common procedures to match buy and sell orders. 
CDA are notably used on large stock markets \shortcite{Tseng2010a}.

@@

Traders are part of a social network where each agent forms two links at random, and then forms links randomly, ensuring that each agent is connected to at least two other agents. 
Every time securities are realized, each trader looks at the performance of her richest neighbor in the network. 
Traders start with the same initial amount of ECU and differences in ECU can only come from market interactions.
Therefore, if some trader is poorer than her richest neighbor, the trader interprets it as a signal that her richest neighbor may have a better model of the climate.
Then, the trader considers adopting the model of her richest neighbor. 
For each trader, the willingness to revise her belief is determined by how ideologically loaded her belief is, which is a parameter we vary in our experiments.

@@

An example of a snapshot of a social network is depicted in Fig.~\ref{fig:social_network}. 
We assume that the initial network is segregated: traders who initially believe in \ce{CO2} forcing (resp. TSI) are more likely to be linked with other traders who also believe in \ce{CO2} (resp. TSI).
This reflects reality where cultural and social groups are highly correlated with climate change beliefs. 
Although traders can change their approximate model over time, the connections between traders do not change as the market unfolds, i.e the edges are fixed. We vary the degree of initial network segregation in our experiments.

@@

  \item Network parameters:
  \begin{itemize}
    \item  \texttt{n.traders}: the number of traders.
    \item  \texttt{n.edg}: the number of edges in the social network.
    \item  \texttt{seg}: the initial degree of homophily in the network. 
    The higher  \texttt{seg}, the higher the initial homophily. 
  \end{itemize}

@@

As a validation that the market is operating correctly, we run the market with actual temperatures 1880 to 2015 and market betting from 1931 to 2014. Before 1980 there is no statistically significant warming trend so we would not expect the market to change beliefs before then. As a global warming signal begins to show up in 1980s, traders begin to converge toward the log(\ce{CO2}) model (Fig.~\ref{fig:time}).

@@

We use the model to perform 10 full simulations for each of the \Sexpr{n_sims} input parameter sets and average the 10 convergence scores. We conducted multiple simulations for the same parameter set because there is stochasticity in the tempertature time series, the social network structure, and the agent decision models. We then conducted a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, and the resulting simulated outcome vector of mean belief convergence scores, $y$ \shortcite{Marino2008,pujol_sensitivity:_2014,saltelli_sensitivity_2009}. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The difference between the partial correlation and the partial rank correlation that we use here is that the outcome variable $y$ is first ranked-transformed in order to allow us capture potentially non-linear relationships: every value, $y_s$, in the data set is replaced by $f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, 
where $y_r$ are the observed values of $y$ in the data set. We conduct 1,000 bootstrapped estimations of the partial rank correlation coefficients to obtain 95\% confidence intervals.

@@

The sensitivity analysis averages over time and thus masks time trends. We randomly drew from the above distributions for ideology, risk tolerance, and the number of traders, and conducted an experiment crossing 95\textsuperscript{th} and 5\textsuperscript{th} percentile values for the number of edges per trader and the segmentation of the social network and both values for the true model. We then collected the time series of convergence across these eight designs for visualization of the distributions of convergence over time (Fig.~\ref{fig:time}). Under most parameterizations, within 15--20 years, the median fraction of traders believing in the true model is over 75\%.

<<time, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Convergence over trading sequences for different degrees of social-network connection and segmentation.", fig.width=10.5, fig.height=8.5, fig.pos='t'>>=
source("utils/time_plot.R")
p1 <- history_box() + theme(text=element_text(family='Palatino'))
p2 <- future_box() + theme(text=element_text(family='Palatino'))
multi_plot(p1, p2, cols=2)
@

@@

The historical sequence in Fig.~\ref{fig:time} shows convergence to the TSI model until the early 1970s, after which traders converge toward the \ce{CO2} model. This is consistent with the fact that greenhouse gas concentrations did not become high enough to begin dominating natural climate variations until the 1970s or so, and it took until the 1990s for the temperature record to show clear signs of anthropogenic interference \shortcite{IPCC_WG1_AR5_2013}.

@@

We simulate two alternative climate futures: one where the natural log of \ce{CO2} is the primary driver of global temperature and one where sun spots are the primary driver. These represent the two most plausible competing views in the public discourse and our analysis is agnostic about which is ``true''.

@@

All code for the model is available at \href{http://johnjnay.com/predMarket/}{johnjnay.com/predMarket/}. This project is a computational test-bed for public policy design: our code can be extended to test the effects that trading strategies, cognitive models, future climate scenarios, and market designs have on the evolution of trader beliefs.

@@

We thank Yevgeniy Vorobeychik for feedback on this project. 
We thank Manuel Naumovich Velasco for sharing the data from his Total Solar Irradiance prediction.
We also thank the authors of \href{http://www.r-project.org}{\textsf{R}} \shortcite{r_core}
and 
\href{http://mc-stan.org}{Stan} \shortcite{Stan2016}.
This manuscript was prepared using 
%RStudio \shortcite{rstudio} and 
knitr \shortcite{xie_knitr_2014}.

>>>>

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public \shortcite{Vandenbergh2013f}. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence \shortcite{Kahan_2011}. The costs of misinformed climate policies are high. If climate change is not human-induced but is believed to be so, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but recognized to be so, the costs of inaction could be devastating. Effective climate policies require acting quickly, so it would be valuable to bring the public to a prompt and accurate consensus on the issue.

@@

\citeN{Bloch_2010} have proposed using derivatives markets to reduce the scientific uncertainty in estimating the impact of future climate change. Existing prediction markets (e.g. \href{https://hypermind.com/hypermind/app.html}{hypermind}, \href{https://www.betfair.com/exchange/}{betfair}, and \href{https://www.predictit.org/}{PredictIt}) focus on near term events such as elections months away, so it is difficult to extrapolate empirical findings to the climate case. Furthermore, we are interested in investigating the unobservable beliefs of traders. Therefore, we turn to simulation modeling informed by climate and economic theory. We simulate a prediction market where traders exchange securities related to climate outcomes to explore whether, and under what social and climate conditions, prediction markets may be useful for increasing convergence of climate beliefs. Our work can be extended as part of a computational design process for effective climate prediction markets.

From a public policy perspective, changing the explanatory models of market participants is one of the most important roles that prediction markets might play. This is perhaps the most important social benefit of prediction markets given the predictive power of statistical supervised learning models, which could possibly provide information at a much lower cost than creating and maintaining a market \shortcite{goel_prediction_2010}. Effective climate policy not only requires an accurate consensus on future climate \emph{outcomes\/}. It also requires an accurate consensus on the causal \emph{mechanisms\/} influencing such outcomes. If people agree that temperature will rise, but some believe it will be due to greenhouse gases, while others believe that it will be caused by increased solar activity, inconsistent and ineffective policies may be implemented.

@@

In our model, traders bet on global temperature anomalies six years in the future. 
During the six-year period, traders buy and sell futures. 
Every year, traders update their models and forecast of future temperatures based on newly available data. 
At the end of each six-year period, winners collect gains and traders revise beliefs about climate models, based on their ideology and the beliefs of top earners in their social network. 
In this section, we describe the models used to generate future temperature data, agents beliefs about those models, the market procedures, and the social network connecting agents. 
We conclude with details on model dynamics and an overview of the model parameters we experimentally vary.

@@

For climate time-series, we use the annual anomaly of global mean temperature. For years from 1880--2014 we use the GISTEMP global mean land-sea annual temperature anomalies \shortcite{GISTEMP_2016,Hansen_2010} and for years from 2015 onward, we project future climates under two alternative theories: In both theories, changes in global temperature are proportional to changes in a deterministic forcing plus a stochastic noise term. For simplicity, we choose two alternative expressions for the deterministic climate forcing: one, which corresponds to conventional climate science, takes the natural logarithm of the atmospheric carbon dioxide concentration in parts per million \shortcite[p.~37]{Archer_2012}, and the other, which corresponds to an alternative theory advocated by many who doubt or reject conventional climate science, takes the total solar irradiance (the brightness of sunlight, in Watts per square meter, at the top of the atmosphere) averaged over the 11-year sunspot cycle \shortcite{Soon_2005}.  Most scientific models of the earth's climate include many forcings, including carbon dioxide, other greenhouse gases, aerosols, total solar irradiance, and more. In these models, the changing $\ce{CO2}$ concentration is, by a large margin, the strongest single forcing \shortcite[p.~14]{IPCC_WG1_AR5_2013}, so for simplicity we consider only \ce{CO2}. Choosing only one forcing term for each of the competing models simplifies comparison because each model has the same number of adjustable parameters.

For $\ce{CO2}$ we used historical emissions through 2005, harmonized with RCP 8.5 representative concentration pathway from 2005 onward \shortcite{rcp_database_2009,Riahi_2011} and for TSI we used the harmonized historical values, with a projection through 2100 from \shortciteN{Velasco_Herrera_2015}, which is, to our knowledge, the only prediction of TSI for the entire 21\textsuperscript{st} century.

Warming coefficients for each model were determined by linear regression of historical temperatures from 1880--2014 against the historical values for each model's forcing term. The noise model was determined by fitting an $\ARMA(p,q)$ model to the residuals from the regression, using the \href{http://mc-stan.org}{Stan} probabilistic modeling language and the \href{https://cran.r-project.org/web/packages/rstan/}{\texttt{rstan}} package \shortcite{Stan2016,RStan2016}. Stan proved more numerically stable than the \Rstat\ \href{http://CRAN.R-project.org/package=nlme}{\texttt{nlme}} package \shortcite{nlme_2016} for fitting ARMA noise models. We identified the optimal model for the auto-correlated noise term by performing the regression analysis for all combinations of $p, q \in [0,1,2]$ and using the Watanabe-Akaike Information Criterion to select the optimal noise model \shortcite{Vehtari_2015}. In both cases (TSI and $\log\ce{CO2}$), the optimal noise model was $\AR(1)$.

Future climates were generated by applying the future climate forcings to Eq.~\ref{eq:climate}:
\begin{equation}
T_{\text{model}}(t) = \beta_{\text{model}} F_{\text{model}}(t) + \varepsilon,
\label{eq:climate}
\end{equation}
where $T_{\text{model}}(t)$ is the temperature at time $t$, under a given model of what causes warming,
$F_{\text{model}}(t)$ is the forcing (TSI or $\log\ce{CO2}$) at time $t$,
$\beta_{\text{model}}$ is the regression coefficient, and
$\varepsilon$ is a noise term.
The coefficient $\beta$ and the parameters of the ARMA noise model were fit to the historical data (1880--2014).
An example realization of future climates for both the $\ln(\ce{CO2})$ and TSI models is shown in Fig.~\ref{fig:climateplot}.

@@

Based on traders' sell and buy orders, traders exchange securities following a continuous-double auction (CDA) procedure (see Model Dynamics below for more details). 
CDAs or some close variants are common procedures to match buy and sell orders. 
CDA are notably used on large stock markets \shortcite{Tseng2010a}.

@@

Traders are part of a social network where each agent forms two links at random, and then forms links randomly, ensuring that each agent is connected to at least two other agents. 
Every time securities are realized, each trader looks at the performance of her richest neighbor in the network. 
Traders start with the same initial amount of ECU and differences in ECU can only come from market interactions.
Therefore, if some trader is poorer than her richest neighbor, the trader interprets it as a signal that her richest neighbor may have a better model of the climate.
Then, the trader considers adopting the model of her richest neighbor. 
For each trader, the willingness to revise her belief is determined by how ideologically loaded her belief is \shortcite{Kahan_2011}, which is a parameter we vary in our experiments.

@@

An example of a snapshot of a social network is depicted in Fig.~\ref{fig:social_network}. 
A segmentation parameter controls the homophily of the network: the extent to which traders are preferentially linked with other traders who share their initial belief in the cause of climate change (\ce{CO2} or TSI), as opposed to traders with the opposite intial belief. This parameter varies from 0 (no preference for like-minded traders) to 1 (traders are only connected to like-minded traders).
%This reflects reality where cultural and social groups are highly correlated with climate change beliefs. 
Although traders can change their beliefs over time, the connections between traders do not change as the market unfolds, i.e the edges are fixed. 
We vary the segmentation parameter in our experiments.

@@

  \item Network parameters:
  \begin{itemize}
    \item  \texttt{n.traders}: the number of traders.
    \item  \texttt{n.edg}: the number of edges in the social network.
    \item  \texttt{seg}: the segmentation parameter for the social network. 
  \end{itemize}

@@

As a validation that the market is operating correctly, we ran the market using actual historical temperatures from 1880 to 2014, with market betting from 1931 to 2014. Greenhouse gas concentrations did not become high enough to begin dominating natural climate variations until the 1970s or so, and it took until the 1990s for the temperature record to show clear signs of anthropogenic interference \shortcite{IPCC_WG1_AR5_2013}. Consistent with this, our simulated historical trading sequence (Fig.~\ref{fig:time}) shows convergence to belief in the TSI model until the early 1970s, after which traders converge toward believing the \ce{CO2} model. 
<<time, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Convergence over trading sequences for different degrees of social-network connection and segmentation.", fig.width=10.5, fig.height=8.5, fig.pos='t'>>=
source("utils/time_plot.R")
p1 <- history_box() + theme(text=element_text(family='Palatino'))
p2 <- future_box() + theme(text=element_text(family='Palatino'))
multi_plot(p1, p2, cols=2)
@

@@

We used the model to perform 10 full simulations for each of the \Sexpr{n_sims} input parameter sets and average the 10 convergence scores. We conducted multiple simulations for the same parameter set because there is stochasticity in the tempertature time series, the social network structure, and the agent decision models. We then conducted a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, and the resulting simulated outcome vector of mean belief convergence scores, $y$ \shortcite{Marino2008,pujol_sensitivity:_2014,saltelli_sensitivity_2009}. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The difference between the partial correlation and the partial rank correlation that we use here is that the outcome variable $y$ is first rank-transformed in order to allow us capture potentially non-linear relationships: every value, $y_s$, in the data set is replaced by $f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, 
where $y_r$ are the observed values of $y$ in the data set. We conduct 1,000 bootstrapped estimations of the partial rank correlation coefficients to obtain 95\% confidence intervals.

@@

The sensitivity analysis averages over time and thus masks time trends. We randomly drew from the above distributions for ideology, risk tolerance, and the number of traders, and conducted an experiment crossing 95\textsuperscript{th} and 5\textsuperscript{th} percentile values for the number of edges per trader and the segmentation of the social network and both values for the true model. We collected the time series of belief-convergence across these eight designs to visualize the distributions of convergence over time (Fig.~\ref{fig:time}). Under most parameterizations, the median fraction of traders believing in the true model reaches 75\% in 10--20 years.

@@

@@

We simulate two alternative climate futures: one where \ce{CO2} is the primary driver of global temperature and one where variations in solar intensity are the primary driver. These represent the two most plausible competing views in the public discourse and our analysis is agnostic about which is ``true''.

@@

All code and data for the model is available at \href{http://johnjnay.com/predMarket/}{johnjnay.com/predMarket/}. This project is a computational test-bed for public policy design: our code can be extended to test the effects that trading strategies, cognitive models, future climate scenarios, and market designs have on the evolution of trader beliefs.

@@

We thank Yevgeniy Vorobeychik for feedback on this project. 
We thank Manuel Naumovich Velasco for sharing the data from his Total Solar Irradiance prediction.
%We also thank the authors of \href{http://www.r-project.org}{\textsf{R}} \shortcite{r_core}
%and 
%\href{http://mc-stan.org}{Stan} \shortcite{Stan2016}.
%This manuscript was prepared using 
%RStudio \shortcite{rstudio} and 
%knitr \shortcite{xie_knitr_2014}.
